% \documentclass[11pt, oneside]{article}      % use "amsart" instead of "article" for AMSLaTeX format


\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{QPL 2021}

\input{preamble}
\input{definitions}

\title{Classifying Complexity with the ZX Calculus}

\author{  Alex Townsend-Teague $^{0,1}$ and Konstantinos Meichanetzidis $^{1,2}$
\institute{$^0$ Mathematical Institute, University of Oxford}
\institute{$^1$ Department of Computer Science, University of Oxford}
\institute{$^2$ Cambridge Quantum Computing Ltd.} }

\def\titlerunning{Classifying Complexity with the ZX Calculus}
\def\authorrunning{Alex Townsend-Teague, Konstantinos Meichanetzidis}


\begin{document}
\maketitle
\begin{abstract}
The ZX calculus is a graphical language which allows for reasoning about suitably represented tensor networks, aka ZX diagrams,
in terms of rewrite rules.
Here, we focus on problems which amount to exactly computing
a scalar encoded as a closed tensor network.
In general, such problems are \#P-hard.
However, there are families of such problems which are known to be in P
when the dimension is bellow a certain value.
By expressing problem-instances from these families as ZX diagrams,
we see that the easy instances belong to the stabiliser fragment of the ZX calculus.
Building on previous work on efficient simplification of qubit stabiliser diagrams, we present simplifying rewrites for the case of qutrits.
Finally, we look at the specific examples of evaluating the Jones polynomial
and of counting graph-colourings.
Our exposition further constitutes the ZX calculus as a suitable and unifying language for studying the complexity of
a broad range of classical or quantum problems.
\end{abstract}

\input{main/intro}
\input{main/qubits}
\input{main/qutrits}
\input{main/case_studies}

\section{Outlook}


In this work, we have advocated for the use of the ZX calculus
as a convenient graphical language, which can treat a broad range of many-body problems on equal footing,
for diagnosing the complexity of interesting problem-families.
When the solution to a problem is encoded as a closed ZX diagram,
this solution can be obtained by full diagram simplification by applying the rules of the calculus.
The problem can be solved efficiently when the diagram encoding its answer is in the stabiliser fragment of the calculus.
In this work specifically, we have fleshed out the simplifications strategies that make this apparent for the case of qutrits.

We also looked at two case studies.
We reviewed complexity results on evaluating the Jones polynomial and counting graph-colourings, problems which can be cast in tensor-network form.
We recover known complexity results, all in the language of ZX calculus.

A main path for future work entails the generalisation
of our stabiliser simplification rules for all prime dimensions.
Tangentially, this also motivates work in circuit extraction \cite{backens2020again} for qudit circuits so that one can attempt graph-theoretic simplification \cite{graph_theoretic_simplification}, analogous to the case of qubits.
Another direction regards defining scalar-exact versions of the rewrite rules which would be necessary for
for concrete applications to problems where $d$-state systems are the native degrees of freedom.
% Finally, it would be interesting to investigate
% the expressive power of generalisations of ZX calculus
% in terms of what problems it can encode.


\section{Acknowledgments}
Alex would like to thank Aleks Kissinger and Quanlong (Harny) Wang for their help throughout this research.
KM wishes to thank Niel de Beaudrap, Aleks Kissinger, Stefanos 
Kourtis, and Quanlong (Harny) Wang for inspiring and helpful discussions.
KM acknowledges financial support from the Royal Commission for the Exhibition of 1851 through a research fellowship.

\bibliographystyle{eptcs}
\bibliography{refs-teague,refs}

\appendix

\input{appendix/zx_calculus}
\input{appendix/pivot}
\section{Spider Elimination Rules for Qutrit ZX}
\input{appendix/eliminate/p_spiders}
\input{appendix/eliminate/n_spiders}
\input{appendix/eliminate/m_spiders}
\input{appendix/pm_boxes}

\end{document} 